Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-07-02 doc/ApTranscriptome_manuscript.Rmd
Final edits for response. Added response letter

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with *enhanced response* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. While temporal patterns of transcriptional activity (e.g. fast vs. slow responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus of the response of transcripts across conditions at the same time point. We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we measured the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression. We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced response hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011]. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim_2014-10-22.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
library(plyr)

@@

Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013]. Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].

We collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A. rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012. These locations are central to each species distributions. Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University). Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation. We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W). We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves [@warren2013].

@@

Ideally, genetically-based variation in gene expression profiles would be identified by comparing individuals completely reared under common-garden conditions to eliminate environmental variation experienced either as adults or during development. However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer. Thus, as is commonly done with other long-lived organisms [@whitehead2006; @barshis2013], we exposed both colonies to common-garden rearing conditions for six months to fully acclimate adult workers to common temperatures. Over this time, roughly 1-2 cohorts of new workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.

Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@cottingham2005]. Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature. To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days. We measured response to temperature with a one-hour static temperature application, which is ecologically-relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012]. Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. Thus, our reactionome characterized early, but not late, responding genes. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, so we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

@@

with to each transcript. Temperature and species were both fixed effects, with a quadratic term included for temperature. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information).

For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression levels to partition transcripts into five expression categories which were defined *a priori*, and could therefore be used to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in one species but did respond in the other. For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C. The two categories which unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).

@@

We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction. For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**. We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species. To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.

To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species. For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts. We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group. For downregulated transcripts (i.e. **Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.

To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A. picea* and for **Low** transcripts in *A. carolinensis*. Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*. In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.

@@

To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, ~80% (7,201) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

>>>>

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  pdf_document:
    fig_caption: true
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
        keep_md: true
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 9% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures, including genes coding for heat shock proteins, than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole-organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such categorical comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often larger than differences in mean values at both the species and the population level [@murren2014].

In this study, we extend the reaction norm approach to RNAseq analysis and introduce the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. Although temporal patterns of transcriptional activity (e.g. fast- vs. slow- responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus here on the response of transcripts across conditions at the same time point.

We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A. picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites. In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*). These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we measured the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression. We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced response hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would activate more genes, and induce them more strongly, in response to low temperatures than would the warm-adapted *A. carolinensis*, which would show greater induction in response to high temperatures.

Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].

Finally, the *genetic assimilation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
library(plyr)

@@

Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in critical maximum and minimum temperatures [@warren2013]. Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].

In fall 2012, we collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A. rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W). These sites are centrally located within each species' geographic range. Species identity was confirmed with morphological characters (Bernice DeMarco, Michigan State University). Colonies of both species were maintained in common conditions at 25°C for 6 months prior to experimentation. Due to colony size limitations, we were unable to determine the critical thermal limits of these particular colonies. In summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W). We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves, following the protocol of Warren & Chick [-@warren2013].

@@

Ideally, genetically-based variation in gene expression profiles would be identified by comparing individuals completely reared under common-garden conditions to eliminate environmental variation experienced either as adults or during development. However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer. Thus, as is commonly done with other long-lived organisms [@whitehead2006; @barshis2013], we exposed both colonies to common-garden rearing conditions for six months to fully acclimate adult workers to common temperatures. Over this time, roughly 1-2 cohorts of new workers are expected to join each colony (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.

Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@cottingham2005]. Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature. To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days. We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012]. Each day, the ants were placed in glass tubes immersed in a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. Thus, our reactionome characterized early, but not late, responding genes. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archaen and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. Because preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

@@

to each transcript. Temperature and species were both fixed effects, with a quadratic term included for temperature. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold. The reaction norms of individual transcripts can be searched by gene ontology and viewed at https://johnsg.shinyapps.io/ApRxN-shinyapp.

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information). For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other. For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the rearing temperature of 25°C. Because expression category was defined by the temperature of maximal expression, both **Low** and **High** categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme. The two categories which could unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).

@@

We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species. To evaluate if shifts from one expression category to another between the two species were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equity.

To test whether the temperature at which thermally-responsive transcripts were activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species, using only the transcripts with a significant species x temperature interaction. For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts. We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group. For downregulated transcripts (i.e. **Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test to compare the mean temperature of down-regulation between species.

To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response variable and the log ratio of relative inducibility as the predictor variable for **High** transcripts in *A. picea* and for **Low** transcripts in *A. carolinensis*. Constitutive expression was defined as predicted expression at 25°C, whereas inducibility of each transcript was defined as *((maximum TPM - minimum TPM) / minimum TPM) x 100*. In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.

@@

To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, after accounting for dependencies among the GO terms.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 9% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. The majority of the responsive transcripts were in the **Low** category for both species (`r Ap.type.table.prop[2,"Low"]`% and `r Ap.type.table.prop[1,"Low"]`% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (`r Ap.type.table.prop[2,"Intermediate"]`%), **High** (`r Ap.type.table.prop[2,"High"]`%),**Bimodal** (`r Ap.type.table.prop[2,"Bimodal"]`%) and **Not Responsive** (`r Ap.type.table.prop[2,"NotResp"]`%) for *A. carolinensis*, and **Bimodal** (`r Ap.type.table.prop[1,"Bimodal"]`%), **High** (`r Ap.type.table.prop[1, "High"]`%), **Intermediate** (`r Ap.type.table.prop[1,"Intermediate"]`%) and **Not Responsive** (`r Ap.type.table.prop[1,"NotResp"]`%) for *A. picea*.
